---
title: "Tutorial Synteny analyses"
author: "Basile PAJOT"
date: "2026-01-27"
output:
  html_document:
    code_folding: hide
    toc: true
    number_sections: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries
```{r}
if (!require("pacman")) install.packages("pacman")
libraries <- c("tidyverse", "ggsci")
pacman::p_load(char = libraries, character.only = TRUE)
if (!require("devtools")) install.packages("devtools")
if (!require("pafr"))devtools::install_github("dwinter/pafr")
library(pafr)

# pacman::p_load(char = c("dplyr", "tibble", "ggplot2", "readr", "tidyr", "stringr", "forcats", "lubridate", "purrr"))
```
# Import useful variables

```{r, class.source="fold-show"}
# Where are located the reference genomes
folder_ref_genomes <- "../Data/Reference_genomes/"
# What are the names of the reference genomes
vector_ref_genomes <- list.files(folder_ref_genomes, pattern = ".fasta$")

# Where to store the output
out_folder <- "../Data/pafs/"
# Create the output folder if it does not exist
if (!dir.exists(out_folder)){
  dir.create(out_folder)
}

# Graph theme
my_theme <- theme_bw() +
  theme(text = element_text(size = 20))
```

# Useful functions

```{r, class.source="fold-show"}
convert_to_numeric_if_possible <- function(x){
  #' Try to convert a column to numeric if it is possible
  #' @description This function tries to convert the values of the specified column
    #' to numeric. If this returns NA (expected for characters or booleans), it 
    #' keeps the column as it is.
  #' @param x This is a column from a dataframe
  #' 
  #' @returns The column as numeric or as it was.
  if (!is.na(as.numeric(x[1]))){
    x <- as.numeric(x)
  }
  return(x)
}

add_table_to_df_in_iteration <- function(df1, df2){
  #' Adds a data frame to another in an iteration
  #' 
  #' In an iteration, if we want to rbind a data frame to another, we have to 
  #' check if the data frame is empty and if it is not, we can rbind the two
  #' data frames together
  #' 
  #' @param df1 (data.frame).
  #'    This data frame is the name of the data frame the data will be added to
  #' @param df2 (data.frame).
  #'    This data frame contains the data to add to the df1.
  #'    
  #' @returns (data.frame).
  #'    This is the binding of the two data frames
  #'
  #' @export
  
  # Then, we have to check if the first data frame is empty
  if (nrow(df1) == 0){
    # If the first data frame is empty, we simply attribute the value of df2 to df1
    df1 <- df2
  }else{
    # Otherwise we rbind the both of them
    df1 <- df1 %>% 
      rbind(df2)
  }
  return(df1)
}

"%!in%" <- function(x, y) (!(x %in% y))

which_not_in <- function(x, y){
  #' Function to get the values that are in x and not in y
  #' @description
    #' This function uses the `which` and the `%!in%` functions to keep the 
    #' values that are in x and not in y.
  #' @param x This vector is the vector to check
  #' @param y This vector is the one being checked
  #' @returns A vector containing the values that are in x and not in y
  x[
   (x %!in% y) %>% 
    which() 
  ] %>% 
    return()
}

source("./Scripts/Synteny_analysis.r")
```

# Import useful data
```{r}
chromosomes_prae_ordered <- read.table("../Data/sumstats_thinned_50k.tsv", header = TRUE) %>% 
  select(Chromosome, Chrom) %>% 
  unique %>% 
  arrange(Chrom)

chromosomes_prae <- import_chromosome_length_from_dict("../Data/Reference_genomes/Jaera_praehirsuta_chromosomes.dict")

chromosomes_ischio <- import_chromosome_length_from_dict("../Data/Reference_genomes/Jaera_ischiosetosa_chromosomes.dict")

chromosomes_albi <- import_chromosome_length_from_dict("../Data/Reference_genomes/Jaera_albifrons_chromosomes.dict")
# Bind the chromosome sizes together
all_chromosomes <- chromosomes_prae %>% 
  mutate(Species = "praehirsuta") %>% 
  rbind(chromosomes_albi %>% 
          mutate(Species = "albifrons")) %>% 
  rbind(chromosomes_ischio %>% 
          mutate(Species = "ischiosetosa")) %>% 
  left_join(chromosomes_prae_ordered, by = "Chromosome") %>% 
  arrange(Species, Chrom, desc(Length)) %>% 
  group_by(Species) %>% 
  mutate(Cumul_position_start = cumsum(lag(Length + 1, default = 0))) %>% 
  ungroup
  
```

# Align reference genomes

To do this, we use `minimap2` that allows to map one genome to another one. It produces `.paf` files. To run minimap2, we start it on the cluster
```{r}
# Do the pairwise alignment of the genomes if it is not yet done
# As the output of minimap2 is a paf file, we simply check if there are any paf files in the output folder
# to check if the reference genomes were aligned.
if (!are_there_paf_files_in_outfolder(out_folder)){
  align_ref_genomes()
}

```

This step takes a bit of time, so you can sit back and eat a snack while it runs on the cluster ;).


# Import the results of the cross alignements
```{r}
albi_on_prae <- import_paf("../Data/pafs/albifrons_aligned_on_praehirsuta.paf")

ischio_on_albi <- import_paf("../Data/pafs/ischiosetosa_aligned_on_albifrons.paf")

# prae_on_ischio <- import_paf("../Data/pafs/praehirsuta_aligned_on_ischiosetosa.paf")
```

```{r}
# # Bind all the tables
# all_correspondences <- prae_on_ischio %>%
#   mutate(query = "praehirsuta",
#          target = "ischiosetosa") %>%
#   rbind(albi_on_prae %>%
#           mutate(query = "albifrons",
#                  target = "praehirsuta")) %>%
#   rbind(ischio_on_albi %>%
#           mutate(query = "ischiosetosa",
#                  target = "albifrons")) %>%
#   mutate(across(contains("name"), ~ str_split_fixed(., "\\.", 2)[, 1]))
all_correspondences <- albi_on_prae %>%
  mutate(query = "albifrons",
         target = "praehirsuta") %>%
  rbind(ischio_on_albi %>%
          mutate(query = "ischiosetosa",
                 target = "albifrons")) %>%
  mutate(across(contains("name"), ~ str_split_fixed(., "\\.", 2)[, 1]))
```
## Plots
### Synteny plots
```{r}
# Get the correspondences between the pairs of chromosomes
correspondence_chroms <- all_correspondences %>% 
  get_corresponding_chromosomes_species()

all_correspondences <- all_correspondences %>% 
  reverse_all_chromosomes(correspondence_chroms)

```

At this point, we have the correspondences between the chromosomes of each species. We need to define a species as a reference for our study. We will use *J. praehirsuta* as reference as it was used in [Ribardière *et al.*, 2025](https://doi.org/10.1101/2025.01.08.631900). This part might need to be changed for your study. If you do not have any already existing reference, the function `get_chromosome_names_all_species` will arrange the chromosomes by decreasing size.
```{r}
#TODO:
# 1. Remove white spacing on top and bottom of plot
# 2. Comment clearly the different functions
# 3. Redact the tutorial to produce a nice document.
# 6. See if there is an optimal way to do the minimum number of permutations to re-arrange the order of the chromosomes using the relative position of species (a lot more complex)
```

```{r}
# The reference is defined here:
reference_species_chromosomes <- all_chromosomes %>% 
  filter(Species == "praehirsuta") %>% 
  select(Chromosome, Chrom, Species) %>% 
  rename(Chromosome_name = Chrom) %>% 
  mutate(Chromosome_name = case_when(
    # We also found fusions of LG1 and LG9 with ischio, so we rename these chroms
    Chromosome == "OZ242388" ~ "LG1.a-b",
    Chromosome == "OZ242385" ~ "LG9.a-b",
    # Lets also add the mitochondrial genome
    Chromosome == "OZ242396" ~ "mito",
    TRUE ~ Chromosome_name
  ))


chromosome_names <- get_chromosome_names_all_species(all_chromosomes = all_chromosomes %>% 
                                                      select(Chromosome, Species, Length),
                                                    correspondences_chromosomes = correspondence_chroms,
                                                    all_alignments = all_correspondences,
                                                    reference_chromosomes = reference_species_chromosomes) %>% 
  # Compute the cumulative positions of the chromosomes for each species
  group_by(Species) %>% 
  arrange(Chromosome_name) %>% 
  mutate(Cumul_position_start_chromosome = cumsum(lag(Length + 1, default = 0))) %>% 
  ungroup %>% 
  mutate(Center_chrom = round(Cumul_position_start_chromosome + Length / 2))



all_correspondences %>% 
  filter(query != "praehirsuta") %>% 
  make_synteny_plot(alignments = .,
                    name_of_chromosomes = chromosome_names %>%
                      filter(Chromosome_name != "mito"),
                    file_name = "Figures/Final_figure_rectangles.png")

all_correspondences %>% 
  filter(query != "praehirsuta") %>% 
  make_synteny_plot(alignments = .,
                    name_of_chromosomes = chromosome_names,
                    file_name = "Figures/mitochrondria.png",
                    only_chromosome = "mito",
                    small_alignment_threshold = 0)
```
# Bin
### Dotplot
```{r}
# Prae on ischio
# prae_on_ischio %>% 
#   mutate(across(contains("name"), ~ str_split_fixed(., "\\.", 2)[, 1])) %>% 
#   select(contains("name"), qstart, qend, tstart, tend) %>% 
#   left_join(all_chromosomes %>% 
#               filter(Species == "praehirsuta") %>% 
#               select(Chromosome, Chrom, Cumul_position_start),
#             join_by("qname" == "Chromosome")) %>% 
#   group_by(qname) %>% 
#   mutate(across(c(qstart, qend), ~ . + Cumul_position_start)) %>% 
#   ungroup %>% 
#   arrange(Chrom) %>% 
#   select(-Cumul_position_start) %>% 
#   left_join(all_chromosomes %>% 
#               filter(Species == "ischiosetosa") %>% 
#               select(Chromosome, Cumul_position_start),
#             by = join_by("tname" == "Chromosome")) %>% 
#   group_by(tname) %>% 
#   mutate(across(c(tstart, tend), ~ . + Cumul_position_start)) %>% 
#   ungroup %>% 
#   ggplot(aes(x = tstart, y = qstart, color = Chrom)) +
#   geom_segment(aes(x = tstart, xend = tend, y = qstart, yend = qend), lwd = 1.5) +
#   geom_vline(data = all_chromosomes %>% 
#                filter(Species == "ischiosetosa"),
#              aes(xintercept = Cumul_position_start)) +
#   geom_hline(data = all_chromosomes %>% 
#                filter(Species == "praehirsuta"),
#              aes(yintercept = Cumul_position_start)) +
#   scale_color_manual(name = "Chromosome",
#                      values = rainbow(11)) +
#   labs(x = "J. ischiosetosa",
#        y = "J. praehirsuta") +
#   guides(colour = guide_legend(override.aes = list(alpha = 1))) +
#   my_theme
# 
# albi_on_prae %>% 
#   mutate(across(contains("name"), ~ str_split_fixed(., "\\.", 2)[, 1])) %>% 
#   select(contains("name"), qstart, qend, tstart, tend) %>% 
#   left_join(all_chromosomes %>% 
#               filter(Species == "albifrons") %>% 
#               select(Chromosome, Cumul_position_start),
#             join_by("qname" == "Chromosome")) %>% 
#   group_by(qname) %>% 
#   mutate(across(c(qstart, qend), ~ . + Cumul_position_start)) %>% 
#   ungroup %>% 
#   select(-Cumul_position_start) %>% 
#   left_join(all_chromosomes %>% 
#               filter(Species == "praehirsuta") %>% 
#               select(Chromosome, Chrom, Cumul_position_start),
#             by = join_by("tname" == "Chromosome")) %>% 
#   group_by(tname) %>% 
#   mutate(across(c(tstart, tend), ~ . + Cumul_position_start)) %>% 
#   ungroup %>% 
#   ggplot(aes(x = tstart, y = qstart, color = Chrom)) +
#   geom_segment(aes(x = tstart, xend = tend, y = qstart, yend = qend), lwd = 1.5) +
#   geom_vline(data = all_chromosomes %>% 
#                filter(Species == "praehirsuta"),
#              aes(xintercept = Cumul_position_start)) +
#   geom_hline(data = all_chromosomes %>% 
#                filter(Species == "albifrons"),
#              aes(yintercept = Cumul_position_start)) +
#   scale_color_manual(name = "Chromosome",
#                      values = rainbow(11)) +
#   labs(x = "J. praehirsuta",
#        y = "J. albifrons") +
#   guides(colour = guide_legend(override.aes = list(alpha = 1))) +
#   my_theme
# 
# ischio_on_albi %>% 
#   mutate(across(contains("name"), ~ str_split_fixed(., "\\.", 2)[, 1])) %>% 
#   select(contains("name"), qstart, qend, tstart, tend) %>% 
#   left_join(all_chromosomes %>% 
#               filter(Species == "ischiosetosa") %>% 
#               select(Chromosome, Cumul_position_start),
#             join_by("qname" == "Chromosome")) %>% 
#   group_by(qname) %>% 
#   mutate(across(c(qstart, qend), ~ . + Cumul_position_start)) %>% 
#   ungroup %>% 
#   select(-Cumul_position_start) %>% 
#   left_join(all_chromosomes %>% 
#               filter(Species == "albifrons") %>% 
#               select(Chromosome, Cumul_position_start),
#             by = join_by("tname" == "Chromosome")) %>% 
#   group_by(tname) %>% 
#   mutate(across(c(tstart, tend), ~ . + Cumul_position_start)) %>% 
#   ungroup %>% 
#   ggplot(aes(x = tstart, y = qstart, color = tname)) +
#   geom_segment(aes(x = tstart, xend = tend, y = qstart, yend = qend), lwd = 1.5) +
#   geom_vline(data = all_chromosomes %>% 
#                filter(Species == "albifrons"),
#              aes(xintercept = Cumul_position_start)) +
#   geom_hline(data = all_chromosomes %>% 
#                filter(Species == "ischiosetosa"),
#              aes(yintercept = Cumul_position_start)) +
#   scale_color_manual(name = "Chromosomes\nalbi",
#                      values = rainbow(12)) +
#   labs(x = "J. albifrons",
#        y = "J. ischiosetosa") +
#   guides(colour = guide_legend(override.aes = list(alpha = 1))) +
#   my_theme
```


```{r}

```

