---
title: "Tutorial Synteny analyses"
author: "Basile PAJOT"
date: "2026-01-27"
output:
  html_document:
    code_folding: hide
    toc: true
    number_sections: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries
```{r}
if (!require("pacman")) install.packages("pacman")
libraries <- c("tidyverse", "pafr", "ggsci")
pacman::p_load(char = libraries, character.only = TRUE)
# pacman::p_load(char = c("dplyr", "tibble", "ggplot2", "readr", "tidyr", "stringr", "forcats", "lubridate", "purrr"))
```
# Import useful variables

```{r, class.source="fold-show"}
# Where are located the reference genomes
folder_ref_genomes <- "/shared/projects/sexisol/archive/genomes_reference/"
# What are the names of the reference genomes
vector_ref_genomes <- list.files(folder_ref_genomes, pattern = ".fasta$")

# Where to store the output
out_folder <- "/shared/projects/sexisol/finalresult/Synteny_analyses/"
# Create the output folder if it does not exist
if (!dir.exists(out_folder)){
  dir.create(out_folder)
}

# Graph theme
my_theme <- theme_bw() +
  theme(text = element_text(size = 20))
```

# Useful functions

```{r, class.source="fold-show"}
get_species_name_from_ref_genome <- function(ref_name, position_in_name=2){
  #' Get the name of the species using the name of the reference genome
  #' @description
    #' This function splits the standardised name of the reference genome to get the name of the species.
    #' The standard name should look something like: `Jaera_albifrons_chromosomes.fasta`.
  #' @param ref_name This is a string with the name of the reference genome. It is like the example in the description.
  #' @param position_in_name `default=2`. This argument describes the position of the name to return in the name of the
  #' reference genome.
  #'
  #' @returns string. It returns the name of the species that was stored in the name of the reference genome.
  return(str_split_1(ref_name, "_")[position_in_name])
}

align_ref_genomes <- function(){
  #' Align reference genomes
  #' @description This way too long function is used to align reference genomes on one another.
    #' It iterates on all unique combination of reference genomes to algin them on one another. A part 
    #' of the function is taylored to my study as I am looking to align reference genomes in a specific
    #' order. The lines can be commented if it is not useful and these lines will be indicated in commentry.
    #' This function calls the minimap2 aligner using a bash script to run it in parallel for all the 
    #' combinations of reference genomes.
  #'
  #' @param None
  #'
  #' @returns None

  # Make the unique pairs of reference genomes
  combination_of_genomes <- combn(vector_ref_genomes, 2)
  # Iterate over these pairs of reference genomes
  for (col in 1:ncol(combination_of_genomes)){
    # Get the names of the reference genomes
    ref_genome_1 <- combination_of_genomes[1, col]
    ref_genome_2 <- combination_of_genomes[2, col]
    # Check that the firt genome is either praehirsuta or albifrons. Otherwise, change them
    # These lines can be commented as they are taylored to my study. You can comment from here ...
    if ((grepl("praehirsuta", ref_genome_2)) & (grepl("albifrons", ref_genome_1))){
      .temp_var. <- ref_genome_2
      ref_genome_2 <- ref_genome_1
      ref_genome_1 <- .temp_var.
      rm(.temp_var.)
    }
    # ... to here
    # Get the name of the output file
    name_output <- paste0(
      out_folder,
      get_species_name_from_ref_genome(ref_genome_2),
      "_aligned_on_",
      get_species_name_from_ref_genome(ref_genome_1),
      ".paf"
    )
    
    # Build the command line
    command_line  <- paste0("/shared/projects/sexisol/script/Basile/Cross_alignment_of_ref_genomes/Scripts/run_minimap2_for_reference_genomes.sh ",
            folder_ref_genomes, ref_genome_1, " ", folder_ref_genomes, ref_genome_2, " ",
            name_output)

    print(paste("command_line: sbatch", command_line))
    
    # Run minimap2 using the bash script on a cluster
    system2(command = "sbatch",
            args = paste0("/shared/projects/sexisol/script/Basile/Cross_alignment_of_ref_genomes/Scripts/run_minimap2_for_reference_genomes.sh ",
            folder_ref_genomes, ref_genome_1, " ", folder_ref_genomes, ref_genome_2, " ",
            name_output))
  }
}

are_there_paf_files_in_outfolder <- function(folder_to_check){
  #' Check if there are any paf files in a folder
  #' @description This function simply checks if there are any `.paf` files in the
    #' specified folder.
  #'
  #' @param folder_to_check This string is the path of the folder to check for `.paf files`
  #'
  #' @returns TRUE/FALSE The function returns TRUE if there is at least one `.paf` file in the
  #' specified folder

  number_paf_files_output <- length(list.files(folder_to_check, patt = ".paf"))
  return(number_paf_files_output > 0)
}

convert_to_numeric_if_possible <- function(x){
  if (!is.na(as.numeric(x[1]))){
    x <- as.numeric(x)
  }
  return(x)
}

import_chromosome_length_from_dict <- function(path){
  read.table(path, skip = 1) %>% 
    select(V2, V3) %>% 
    rename(Chromosome = V2,
           Length = V3) %>% 
    mutate(across(everything(), ~ str_split_fixed(., ":", 2)[, 2]),
           Length = as.numeric(Length),
           Chromosome = str_split_fixed(Chromosome, "\\.", 2)[, 1]) %>% 
    return()
}

import_paf <- function(path){
  read_paf(path) %>% 
  plyr::ldply() %>% 
  column_to_rownames(".id") %>% 
  t %>% 
  as_tibble() %>% 
  mutate(across(everything(), ~ suppressWarnings(convert_to_numeric_if_possible(.)))) %>% 
    return()
}
```

# Import useful data
```{r}
chromosomes_prae_ordered <- read.table("/shared/projects/sexisol/finalresult/ddRAD_multiple_choice_exp/09_thin_vcf/Information_vcf/sumstats_thinned_50k.tsv", header = TRUE) %>% 
  select(Chromosome, Chrom) %>% 
  unique %>% 
  arrange(Chrom)

chromosomes_prae <- import_chromosome_length_from_dict("/shared/projects/sexisol/archive/genomes_reference/Jaera_praehirsuta_chromosomes.dict")

chromosomes_ischio <- import_chromosome_length_from_dict("/shared/projects/sexisol/archive/genomes_reference/Jaera_ischiosetosa_chromosomes.dict")

chromosomes_albi <- import_chromosome_length_from_dict("/shared/projects/sexisol/archive/genomes_reference/Jaera_albifrons_chromosomes.dict")
# Bind the chromosome sizes together
all_chromosomes <- chromosomes_prae %>% 
  mutate(Species = "praehirsuta") %>% 
  rbind(chromosomes_albi %>% 
          mutate(Species = "albifrons")) %>% 
  rbind(chromosomes_ischio %>% 
          mutate(Species = "ischiosetosa")) %>% 
  left_join(chromosomes_prae_ordered, by = "Chromosome") %>% 
  arrange(Species, Chrom, desc(Length)) %>% 
  group_by(Species) %>% 
  mutate(Cumul_position_start = cumsum(lag(Length + 1, default = 0))) %>% 
  ungroup
  
```

# Align reference genomes

To do this, we use `minimap2` that allows to map one genome to another one. It produces `.paf` files. To run minimap2, we start it on the cluster
```{r}
# Do the pairwise alignment of the genomes if it is not yet done
# As the output of minimap2 is a paf file, we simply check if there are any paf files in the output folder
# to check if the reference genomes were aligned.
if (!are_there_paf_files_in_outfolder(out_folder)){
  align_ref_genomes()
}

```

This step takes a bit of time, so you can sit back and eat a snack while it runs on the cluster ;).


# Test of Asynt
```{r}
# Download and load the library
source("Scripts/Asynt.R")
```

To use `syntenyPlotteR`, we have to reformat the data type as follows

```{r, class.source="fold-show"}
# Import the paf files
prae_on_ischio <- import.paf("/shared/projects/sexisol/finalresult/Synteny_analyses/praehirsuta_aligned_on_ischiosetosa.paf")
prae_data <- import.genome("/shared/projects/sexisol/archive/genomes_reference/Jaera_praehirsuta_chromosomes.fasta.fai")
ischio_data <- import.genome("/shared/projects/sexisol/archive/genomes_reference/Jaera_ischiosetosa_chromosomes.fasta.fai")

# Filter to remove short alignments (Rlen and Qlen give the number of bases on the reference and query that are included in an alignment tract)
alignments <- prae_on_ischio %>% 
  filter(Rlen >= 200,
         Qlen >= 200)

# Filter to remove large chromosomes
get_list_of_large_chromosomes <- function(list_ref_genome, limit_size){
  names(list_ref_genome$seq_len)[which(list_ref_genome$seq_len >= limit_size)] %>%
    return()
}
large_chromosomes_prae <- get_list_of_large_chromosomes(prae_data, 1000000)
large_chromosomes_ischio <- get_list_of_large_chromosomes(ischio_data, 1000000)

alignments <- alignments %>% 
  filter(query %in% large_chromosomes_prae & reference %in% large_chromosomes_ischio)

# Filter to keep only the scaffolds that share a loarge proportion of aligned sequence with the reference scaffold
query_aln_len <- get.query.aln.len(alignments) %>%
  as.data.frame() %>% 
  rename(alignment_length = ".") %>% 
  rownames_to_column("Chr")

# Look at the distribution of total alignment length for each scaffold
(query_aln_len %>% 
  ggplot(aes(x = Chr, y = alignment_length)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45))) %>% 
  ggsave(plot = ., "~/Rplot.png", scale = 5, units = "px", width = 500, height = 500)

# A similar approach can be used with aligned proportion, which may be more appropriate if query scaffolds are very variable in length
query_aln_prop <- get.query.aln.prop(alignments, query_lens = prae_data$seq_len) %>% 
  as.data.frame() %>% 
  rename(alignment_prop = ".") %>% 
  rownames_to_column("Chr")
(query_aln_prop %>% 
  ggplot(aes(x = Chr, y = alignment_prop)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45))) %>% 
  ggsave(plot = ., "~/Rplot.png", scale = 5, units = "px", width = 500, height = 500)

# Filter the alignments for only those contigs with large proportion aligned to the reference contig of interest
alignments <- query_aln_prop  %>% 
  filter(alignment_prop > 0.1) %>% 
  select(-alignment_prop) %>% 
  rename(query = Chr) %>% 
  left_join(alignments, by = "query")


synblocks <- get.synteny.blocks.multi(alignments, min_subblock_size=200) %>% 
  get.synteny.blocks.multi(min_subblock_size=2000) %>% 
  get.synteny.blocks.multi(min_subblock_size=20000)

  

#and finally we can visualise the alignemnt (see more examples of visualisation below)
png(filename = "/shared/projects/sexisol/finalresult/Synteny_analyses/Figures/prae_on_ischio.png")
plot.alignments.multi(synblocks, reference_lens=ischio_data$seq_len, query_lens=prae_data$seq_len, sigmoid=T)
dev.off()

```

This is a good idea. You can also try with the `synblocs` to get less points. It seems that you need to find a trade-off between the number of points and the number of "parasite points". 
In addition, you can do like in a manhattan plot with LG1-LGn and get some cumulative positions.
The asyntR package does not seem to be very useful here (except for the `get.synteny.blocks.multi` function. See if there is a way to do a ggplot of the synteny plots).
```{r}
(alignments %>% 
  ggplot(aes(x = Rstart, y = Qstart, colour = query)) +
  geom_point(alpha = 0.5) +
  facet_wrap(vars(reference), scales = "free") +
  labs(x = "ischio", y = "prae")+
  theme(text = element_text(size = 20))) %>% 
  ggsave(plot = ., "/shared/projects/sexisol/finalresult/Synteny_analyses/Figures/prae_on_ischio_dotplot.png",
         height = 700, width = 700, scale = 10, units = "px")
```


# Test pafr
```{r}
prae_on_ischio <- import_paf("/shared/projects/sexisol/finalresult/Synteny_analyses/praehirsuta_aligned_on_ischiosetosa.paf")

albi_on_prae <- import_paf("/shared/projects/sexisol/finalresult/Synteny_analyses/albifrons_aligned_on_praehirsuta.paf")

ischio_on_albi <- import_paf("/shared/projects/sexisol/finalresult/Synteny_analyses/ischiosetosa_aligned_on_albifrons.paf")

# Bind all the tables
all_correspondances <- prae_on_ischio %>% 
  mutate(query = "praehirsuta",
         target = "ischiosetosa") %>% 
  rbind(albi_on_prae %>% 
          mutate(query = "albifrons",
                 target = "praehirsuta")) %>% 
  rbind(ischio_on_albi %>% 
          mutate(query = "ischiosetosa",
                 target = "albifrons")) %>% 
  mutate(across(contains("name"), ~ str_split_fixed(., "\\.", 2)[, 1]))
```
## Plots
### Dotplot
```{r}
# Prae on ischio
prae_on_ischio %>% 
  mutate(across(contains("name"), ~ str_split_fixed(., "\\.", 2)[, 1])) %>% 
  select(contains("name"), qstart, qend, tstart, tend) %>% 
  left_join(all_chromosomes %>% 
              filter(Species == "praehirsuta") %>% 
              select(Chromosome, Chrom, Cumul_position_start),
            join_by("qname" == "Chromosome")) %>% 
  group_by(qname) %>% 
  mutate(across(c(qstart, qend), ~ . + Cumul_position_start)) %>% 
  ungroup %>% 
  arrange(Chrom) %>% 
  select(-Cumul_position_start) %>% 
  left_join(all_chromosomes %>% 
              filter(Species == "ischiosetosa") %>% 
              select(Chromosome, Cumul_position_start),
            by = join_by("tname" == "Chromosome")) %>% 
  group_by(tname) %>% 
  mutate(across(c(tstart, tend), ~ . + Cumul_position_start)) %>% 
  ungroup %>% 
  ggplot(aes(x = tstart, y = qstart, color = Chrom)) +
  geom_segment(aes(x = tstart, xend = tend, y = qstart, yend = qend), lwd = 1.5) +
  geom_vline(data = all_chromosomes %>% 
               filter(Species == "ischiosetosa"),
             aes(xintercept = Cumul_position_start)) +
  geom_hline(data = all_chromosomes %>% 
               filter(Species == "praehirsuta"),
             aes(yintercept = Cumul_position_start)) +
  scale_color_manual(name = "Chromosome",
                     values = rainbow(11)) +
  labs(x = "J. ischiosetosa",
       y = "J. praehirsuta") +
  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
  my_theme

albi_on_prae %>% 
  mutate(across(contains("name"), ~ str_split_fixed(., "\\.", 2)[, 1])) %>% 
  select(contains("name"), qstart, qend, tstart, tend) %>% 
  left_join(all_chromosomes %>% 
              filter(Species == "albifrons") %>% 
              select(Chromosome, Cumul_position_start),
            join_by("qname" == "Chromosome")) %>% 
  group_by(qname) %>% 
  mutate(across(c(qstart, qend), ~ . + Cumul_position_start)) %>% 
  ungroup %>% 
  select(-Cumul_position_start) %>% 
  left_join(all_chromosomes %>% 
              filter(Species == "praehirsuta") %>% 
              select(Chromosome, Chrom, Cumul_position_start),
            by = join_by("tname" == "Chromosome")) %>% 
  group_by(tname) %>% 
  mutate(across(c(tstart, tend), ~ . + Cumul_position_start)) %>% 
  ungroup %>% 
  ggplot(aes(x = tstart, y = qstart, color = Chrom)) +
  geom_segment(aes(x = tstart, xend = tend, y = qstart, yend = qend), lwd = 1.5) +
  geom_vline(data = all_chromosomes %>% 
               filter(Species == "praehirsuta"),
             aes(xintercept = Cumul_position_start)) +
  geom_hline(data = all_chromosomes %>% 
               filter(Species == "albifrons"),
             aes(yintercept = Cumul_position_start)) +
  scale_color_manual(name = "Chromosome",
                     values = rainbow(11)) +
  labs(x = "J. praehirsuta",
       y = "J. albifrons") +
  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
  my_theme

ischio_on_albi %>% 
  mutate(across(contains("name"), ~ str_split_fixed(., "\\.", 2)[, 1])) %>% 
  select(contains("name"), qstart, qend, tstart, tend) %>% 
  left_join(all_chromosomes %>% 
              filter(Species == "ischiosetosa") %>% 
              select(Chromosome, Cumul_position_start),
            join_by("qname" == "Chromosome")) %>% 
  group_by(qname) %>% 
  mutate(across(c(qstart, qend), ~ . + Cumul_position_start)) %>% 
  ungroup %>% 
  select(-Cumul_position_start) %>% 
  left_join(all_chromosomes %>% 
              filter(Species == "albifrons") %>% 
              select(Chromosome, Cumul_position_start),
            by = join_by("tname" == "Chromosome")) %>% 
  group_by(tname) %>% 
  mutate(across(c(tstart, tend), ~ . + Cumul_position_start)) %>% 
  ungroup %>% 
  ggplot(aes(x = tstart, y = qstart, color = tname)) +
  geom_segment(aes(x = tstart, xend = tend, y = qstart, yend = qend), lwd = 1.5) +
  geom_vline(data = all_chromosomes %>% 
               filter(Species == "albifrons"),
             aes(xintercept = Cumul_position_start)) +
  geom_hline(data = all_chromosomes %>% 
               filter(Species == "ischiosetosa"),
             aes(yintercept = Cumul_position_start)) +
  scale_color_manual(name = "Chromosomes\nalbi",
                     values = rainbow(12)) +
  labs(x = "J. albifrons",
       y = "J. ischiosetosa") +
  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
  my_theme
```
### Synteny plots
```{r}
aln_prae_ischio <- read_paf("/shared/projects/sexisol/finalresult/Synteny_analyses/praehirsuta_aligned_on_ischiosetosa.paf") %>% 
  subset(alen > 5e5)

plot_synteny(aln_prae_ischio, q_chrom = "OZ242391.1", t_chrom = "OZ239486.1", centre = T)

plot_coverage(aln_prae_ischio, fill="qname")
```
### Manual plots
```{r}
correspondance_chroms <- all_correspondances %>% 
  group_by(query, target, qname, tname) %>% 
  summarise(total_length = sum(alen)) %>% 
  arrange(desc(total_length)) %>% 
  dplyr::slice(1) %>% 
  ungroup() %>% 
  select(-total_length)


albi_chroms <- correspondance_chroms %>% 
  filter(query == "albifrons") %>% 
  left_join(all_chromosomes %>% 
              select(Chromosome, Chrom),
            by = join_by("tname" == "Chromosome")) %>% 
  mutate(Chrom = case_when(
    Chrom == "LG6.a" ~ "LG6.a-b",
    Chrom == "LG5.b-c" & tname == "OZ349670" ~ "LG5.b",
    Chrom == "LG5.b-c" & qname == "OZ349669" ~ "LG5.c",
    qname == "OZ349663" ~ "LG8",
    TRUE ~ Chrom
  ))

ischio_chroms <- correspondance_chroms %>% 
  filter(target == "ischiosetosa") %>% 
  left_join(all_chromosomes %>% 
              select(Chromosome, Chrom),
            by = join_by("qname" == "Chromosome"))

# Redo the positions of the chromosomes
all_chromosomes <- albi_chroms %>%
  rename(Species = query,
         Chromosome = qname) %>% 
  select(Species, Chromosome, Chrom) %>% 
  rbind(chromosomes_prae_ordered %>% 
          mutate(Species = "praehirsuta")) %>% 
  rbind(ischio_chroms %>% 
          rename(Species = target,
                 Chromosome = tname) %>% 
          select(Species, Chromosome, Chrom)) %>% 
  left_join(all_chromosomes %>% 
              select(Chromosome, Species, Length),
            by = c("Chromosome", "Species")) %>% 
  group_by(Species) %>% 
  arrange(Species, Chrom) %>% 
  mutate(Position_cumul = cumsum(lag(Length + 1, default = 0))) %>% 
  ungroup



# Test just for one chromosome pair
correspondance_chroms %>% 
  filter(query != "praehirsuta") %>%
  left_join(all_correspondances,
            by = c("query", "target", "qname", "tname")) %>% 
  left_join(all_chromosomes %>% 
              select(Chromosome, Species, Position_cumul),
            by = join_by("qname" == "Chromosome", "query" == "Species")) %>% 
  mutate(qstart = qstart + Position_cumul) %>% 
  select(-Position_cumul) %>% 
  left_join(all_chromosomes %>% 
              select(Chromosome, Species, Position_cumul),
            by = join_by("tname" == "Chromosome", "target" == "Species")) %>% 
  mutate(tstart = tstart + Position_cumul) %>% 
  filter(alen > 1e4) %>% 
  select(query, target, qname, tname, qstart, tstart) %>% 
  pivot_longer(cols = contains("name"), names_to = "tata", values_to = "Chromosome") %>% 
  mutate(query = factor(query, levels = c("ischiosetosa", "albifrons", "praehirsuta"))) %>% 
  left_join(all_chromosomes %>% 
              select(Chromosome, Chrom),
            by = "Chromosome") %>% 
  mutate(Chrom = str_split_fixed(Chrom, "\\.", 2)[, 1]) %>% 
  ggplot() +
  geom_segment(aes(x = qstart, xend = tstart, y = query, yend = target, colour = Chrom), alpha = 0.02) +
  my_theme +
  theme(legend.position = "none")



correspondance_chroms %>% 
  filter(query == "albifrons") %>% 
  left_join(all_chromosomes %>% 
              select(Chromosome, Chrom),
            by = join_by("tname" == "Chromosome")) %>% 
  arrange(Chrom) %>% 
  left_join(all_correspondances,
            by = c("qname", "tname", "query", "target")) %>% 
  group_by(qname) %>% 
  mutate()

  
```


# Test SyntenyploR
```{r}
pacman::p_load("syntenyPlotteR")
reformat.syntenyData(file_data = "/shared/projects/sexisol/finalresult/Synteny_analyses/praehirsuta_aligned_on_ischiosetosa.paf",
                     directory = "/shared/projects/sexisol/finalresult/Synteny_analyses/test/",
                     filename = "prae_on_ischio",
                     reference.species = "ischio",
                     target.species = "prae")

draw.eh(directory = "/shared/projects/sexisol/finalresult/Synteny_analyses/test/",
        output = "draw_eh", chrRange = 1:12,
        data_file = "/shared/projects/sexisol/finalresult/Synteny_analyses/test/prae_on_ischio.txt")

```

This package does not work, at least the reformatting part from a paf file. The reformatting must be done from a DESCHRAMBLER format or a inferCARs format.

